Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-12-26 paper/main.tex
reformat keywords
rewrite abstract of paper

\textbf{Discussed Keywords (Working Title)}
\begin{itemize}
    \item Parameter Estimation
    \begin{itemize}
        \item Mechanical Properties
        \item Rod-shaped?
        \item Single-cell
    \end{itemize}
    \item \ac{abm}
    \item Methods
\end{itemize}
% keywords can be removed
% \keywords{}

%###################################################################################################
\section*{Abstract}
Elongated bacteria such as \acl{ecoli} and \acl{bsubtilis} possess distinct mechanical
and physiological properties that shape their growth, division, and collective behaviors.
Despite the importance of these organisms, existing computational frameworks rarely capture their
flexible filamentous morphology, limiting the ability to study their mechanics in a systematic and
quantitatively grounded way.
Agent-based models are well suited for representing individual cells and their interactions, yet
their application to flexible rod-shaped bacteria has been hindered by the absence of appropriate
mechanical formulations and by the lack of practical methods for estimating model parameters from
experimental data.
In this work, we introduce a computational model that represents elongated bacteria as deformable
rods capable of growth, elastic bending, and mechanical interactions.
We develop a set of imaging-based and geometry-based comparison methods that extract cell shape and
position information from segmented microscopy data and quantitatively compare it with model
simulations at the single-cell level.
These methods enable the estimation of model parameters across multiple experimental scenarios of
increasing complexity, including cases that exhibit substantial cell-to-cell variability.
Our results demonstrate that parameter estimation for complex agent-based models of flexible
bacteria is feasible and can yield biologically interpretable constraints on cellular mechanics.
The modeling and analysis framework is modular, generalizable, and readily adaptable to a broad
range of rod-shaped species.
By integrating our mechanical model into an extensible computational platform, we provide tools that
support more rigorous model construction, systematic calibration, and mechanistic interpretation.
This work establishes a foundation for quantitatively studying the mechanical behavior of
filamentous bacteria and paves the way for future applications ranging from biofilm formation to
multicellular
patterning.

>>>>

\textbf{Discussed Keywords (Working Title)}\\
Parameter Estimation (Mechanical Properties, Rod-shaped, Single-cell), \ac{abm}, Methods
% keywords can be removed
% \keywords{}

%###################################################################################################
\section*{Abstract}
Elongated bacteria such as \acl{ecoli} and \acl{bsubtilis} possess distinct mechanical
and physiological properties that shape their growth, division, and collective behaviors.
Despite the importance of these organisms, existing computational frameworks rarely capture their
flexible filamentous morphology, limiting the ability to study their mechanics in a systematic and
quantitatively grounded way.
Agent-based models are well suited for representing individual cells and their interactions, yet
their application to flexible rod-shaped bacteria has been hindered by the absence of appropriate
mechanical formulations and by the lack of practical methods for estimating model parameters from
experimental data.
In this work, we introduce a computational model that represents elongated bacteria as deformable
rods capable of growth, elastic bending, and mechanical interactions.
We develop a novel set of imaging- and geometry-based methods that extract cell shape and positional
information from segmented microscopy data and quantitatively compare it with model simulations at
the single-cell level.
These methods allow for a detailed analysis of the underlying dynamics of the model.
They allow us to apply maximum-likelihood and profile-likelihood estimation techniques to determine
the parameters of the underlying model.
We apply them across multiple experimental scenarios of increasing complexity, including cases that
exhibit substantial cell-cell variability and show how these methods can be used to perform model
validation and reduction.
Our results demonstrate that direct parameter estimation with cellular heterogeneity for complex
agent-based models of bacteria is feasible and can yield biologically interpretable constraints on
cellular parameters.
The developed are readiliy available as a modular Python package backed by a high-performing backend
written in Rust and integrate into our previously devleped agent-based simulation framework
\texttt{cellular\_raza}.
